---
title: "EpiFusion Tutorial"
#author: "Your name here"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.align = "center")
```

## Introduction

This R Markdown document will guide you through the process of parsing and interpreting the output of the EpiFusion tutorial. We will use a selection of functions that can also be found in the **EpiFusion_utilities.R** script on the GitHub repository.

The first step of the tutorial is to point the document to the filepath of the output folder made by EpiFusion. We will also source the `EpiFusion_utilities.R` script to load the necessary functions to look through this output. The current filepath assumes the folder is in your working directory; you may need to change this depending on where your output folder was generated.

```{r intro}
outputfolder <- "baseline_combined_example_results/"
```


## Loading EpiFusionUtilities
The next step will be to install or load our companion R package, EpiFusionUtilities, which has functions that parse EpiFusion output into a useful R compatible format.

```{r loadpackage}
devtools::install_github("https://github.com/ciarajudge/EpiFusionUtilities")
library(EpiFusionUtilities)
```

We will also use ggplot in some of our plotting later in the tutorial:
```{r loadggplot}
install.packages("ggplot2")
library(ggplot2)
```

## Loading Raw EpiFusion Results
Here we load the raw results of our EpiFusion analysis into our R session. This function loads the raw posterior samples for each chain and some other information such as the number of chains etc into our session. We use the `load_raw_epifusion()` function for this, which takes a path to an output folder as it's input:

```{r loadrawoutput}
raw_epifusion_output <- load_raw_epifusion(outputfolder)
```

The resulting output is a large list with:
1. The number of chains
2. The number of samples per chain
3. The likelihoods of each chain at each sampled step
4. The acceptance rate of each chain between samples
5. Posterior samples of infection trajectories per chain
6. Posterior samples of rt trajectories per chain
7. Posterior samples of parameter values per chain
8. Simulated case incidence (if it's a combined or epi only analysis)

## Checking convergence

Next we will take a look at the trace plots of the chain likelihood, to check that the model has run nicely. For this we can use the `plot_likelihood_trace` function, which takes the raw epifusion object as its input.

```{r plottraces, echo=FALSE}
plot_likelihood_trace(raw_epifusion_output)
```

Have a look at the likelihood trace for each chain run. They should start in different places but eventually converge to roughly the same value. Keep an eye out for chains getting stuck (staying at the same value for a long time). This does happen occassionally, and we are working on introducing adaptive Metropolis Hastings MCMC sampling to EpiFusion to fix this. Once you have had a look at the trace plot you can choose what proportion of each chain to discard as burn-in. EpiFusion models tend to converge relatively quickly, so the default in this document will be 10%. However, you can edit this as needed.

```{r set_burnin}
burn_in <- 0.1
```

## Extracting your posterior minus burn-in
Now that we have decided on the proportion of samples to discard as burn-in, we can parse the raw output into a more useful format. We do this with the `extract_posterior_epifusion()` function which takes the raw input and a proportion to discard as burn in as its arguments:

```{r extract_posterior}
final_outputs <- extract_posterior_epifusion(raw_epifusion_output, burn_in)
```

This function produces infection and Rt trajectory and parameter posteriors. For the trajectories, the samples (minus the burn-in and with aggregated across chains) are provided, along with mean trajectories and 95%, 88% and 66% HPD intervals. For the parameters, the samples are provided and Rhat and Effective Sample Sizes for each parameter to help assess convergence and mixing. If you have any fixed parameters you will get a warning that says 'no variability observed in a parameter; setting batch size to 0' when you run this function. Don't worry about this - this just means it's not possible to compute an Rhat value for chains with constant equal values. However, if this comes up for an analysis with no fixed parameters, you may have convergence issues - check the likelihood trace to further examine this.

## Examining infection trajectory posteriors
### Trajectories table
Next we'll examine the infection trajectories sampled by your EpiFusion model. First let's take a look at the first sample by selecting it from the final_outputs object.

```{r load_trajectories_table}
unlist(final_outputs$infection_trajectories$infection_trajectory_samples[1,])
```

### Trajectories plot
It can be difficult to get a sense for what the trajectories are really saying in table form, so let's plot them ggplot. We already have the mean and HPD intervals, which makes it easy to put them into a table and plot them. We'll add the Rt trajectories too while we're at it.

```{r plot_trajectories}
trajectorytable <- data.frame(Time = 1:length(final_outputs$infection_trajectories$mean_infection_trajectory),
                        Mean_Infected = final_outputs$infection_trajectories$mean_infection_trajectory,
                        Lower95_Infected = final_outputs$infection_trajectories$infection_trajectory_hpdintervals$HPD0.95$Lower,
                        Upper95_Infected = final_outputs$infection_trajectories$infection_trajectory_hpdintervals$HPD0.95$Upper,
                        Lower88_Infected = final_outputs$infection_trajectories$infection_trajectory_hpdintervals$HPD0.88$Lower,
                        Upper88_Infected = final_outputs$infection_trajectories$infection_trajectory_hpdintervals$HPD0.88$Upper,
                        Lower66_Infected = final_outputs$infection_trajectories$infection_trajectory_hpdintervals$HPD0.66$Lower,
                        Upper66_Infected = final_outputs$infection_trajectories$infection_trajectory_hpdintervals$HPD0.66$Upper,
                        Mean_Rt = final_outputs$rt_trajectories$mean_rt_trajectory,
                        Lower95_Rt = final_outputs$rt_trajectories$rt_trajectory_hpdintervals$HPD0.95$Lower,
                        Upper95_Rt = final_outputs$rt_trajectories$rt_trajectory_hpdintervals$HPD0.95$Upper,
                        Lower88_Rt = final_outputs$rt_trajectories$rt_trajectory_hpdintervals$HPD0.88$Lower,
                        Upper88_Rt = final_outputs$rt_trajectories$rt_trajectory_hpdintervals$HPD0.88$Upper,
                        Lower66_Rt = final_outputs$rt_trajectories$rt_trajectory_hpdintervals$HPD0.66$Lower,
                        Upper66_Rt = final_outputs$rt_trajectories$rt_trajectory_hpdintervals$HPD0.66$Upper)

ggplot(trajectorytable, aes(x = Time)) +
  geom_line(aes(y = Mean_Infected), col = "#2aac6d")+
  geom_ribbon(aes(ymin = Lower95_Infected, ymax = Upper95_Infected), fill = "#2aac6d", alpha = 0.3) +
  geom_ribbon(aes(ymin = Lower88_Infected, ymax = Upper88_Infected), fill = "#2aac6d", alpha = 0.3) +
  geom_ribbon(aes(ymin = Lower66_Infected, ymax = Upper66_Infected), fill = "#2aac6d", alpha = 0.3)

```

The process of making this table is quite arduous - we'll shortly add a function to EpiFusionUtilities that can do it for you!

## Examining R(t) trajectory posteriors
Now let's look at the same plot, but for Rt. We can use the same `trajectorytable` as before, and just slightly adjust the same plotting code:

```{r plot_rttrajectories}
ggplot(trajectorytable, aes(x = Time)) +
  geom_line(aes(y = Mean_Rt), col = "#2aac6d")+
  geom_ribbon(aes(ymin = Lower95_Rt, ymax = Upper95_Rt), fill = "#2aac6d", alpha = 0.3) +
  geom_ribbon(aes(ymin = Lower88_Rt, ymax = Upper88_Rt), fill = "#2aac6d", alpha = 0.3) +
  geom_ribbon(aes(ymin = Lower66_Rt, ymax = Upper66_Rt), fill = "#2aac6d", alpha = 0.3)

```


## Parameter Posteriors
Another interesting thing to look at are the posteriors of the parameters of the MCMC analysis. For example, let's look at the density of the `psi` genomic sampling parameter:

```{r plotpsi}
ggplot(data.frame(psi = final_outputs$parameters$psi$samples), aes(x = psi)) +
  geom_density()
```

Here we coerce the psi samples into a data frame a plot the density with ggplot. We can also get some other statistics for psi:

The Rhat is the gelman-rubin convergence metric, where closer to 1 indicates convergence. The threshold may vary depending on your analysis but it's traditionally recommended to try get this metric below 1.015:

```{r psigr}
final_outputs$parameters$psi$rhat
```

The effective sample size refers to the equivalent number of independent observations that would yield the same statistical precision as the observed data, taking into account the autocorrelation structure introduced by the markov chain. Again the goal for this can differ from analysis to analysis, but you should probably at least aim to get higher than ~100.

```{r ess}
final_outputs$parameters$psi$ess
```

# Conclusion
The above covers the basics of parsing EpiFusion output, and introduces some of the functions of EpiFusionUtilities. We stress that EpiFusion is still in its infancy, with more improvements happening every day, but we hope this tutorial will have peaked your interest! For more information we recommend checking out the EpiFusion wiki.
